---
title: "Statistical Analysis of Immunoglobin Isotype Fingerprints Across Afflictions"
author: "Jakob Gad Spoorendonk"
date: 2026-02-11
header-includes:
  - \supervisors{Veit Stefan Schwämmle\protect\\ Xuan-Tung Trinh}
bibliography: references.bib
format:
  pdf:
    pdf-engine: xelatex
    cite-method: biblatex
    toc: true
    toc-depth: 2
    toc-float: false
    toc-title: "Contents"
    include-in-header:
      - text: |
          \usepackage{tikz}
          \usetikzlibrary{positioning, chains}
          \usepackage{graphicx}
          \graphicspath{{Pictures/}}
      - file : preamble.tex
---
\newpage
# Introduction

## Antibodies  
Antibodies are proteins that make up a part of the adaptive immune system. They are central in humoral immunity attatching themselves to recognized targets to trigger various reactions. All human antibodies are heterodimers that are made up of two dimers of a light and a heavy chain pairing. Both chains contain a variable region. The light chain contains a single conserved region, while the heavy chain contains either 4 (IgM and IgE) or 3 (the rest) conserved regions, this depends on which isotype of antibody it is. Between the first conserved region - which is connected to the variable region - is a hinge region, depending on the isotype the hinge region can be longer or shorter, controlling the flexibility of the antibody arms [@Kretschmer2017]. The different isotypes can interact with different groups of immune receptors, refered to as Fc receptors.

\begin{center}
\includegraphics*[height=10cm]{Antibodies_Kretschmer}
\end{center}
Figure 1: Graphic overview of antibody architecture by Kreschmer et al. 2017 [@Kretschmer2017]  


### Variable Region
The variable region of an antibody is made up of the N-terminal region of the light and heavy chains [@Chi2020]. The variable region is the part of the antibody that react with an antigen to trigger the receptor complex [@Kretschmer2017]. The variability is generated in several ways and allows the body to produce antibodies that can quickly be made specific for a single protein that may appear.


### Constant Region
The constant region of the heavy chain is what determines the isotype of the antibody. The isotype of the antibody decides which effector function the antibody will trigger when activated [@Stavnezer2008]. The function of an antibody is dependent on which immune receptors - usually named Fc_R_ receptors - it can interact with, as it is the conserved part of the antibody that interacts with the receptor, the structure of the conserved part decides the reaction that is triggered. The antibodies and their respective receptors function as the initiation receptor for the produced reaction [@Kretschmer2017]. 


### Hinge Region - NOT DONE YET
The different isotypes and subtypes have different hinge regions. The hinge region of an antibody is found between the first and second constant regions of the heavy chain. The length and disulfide bond of the hinge region affect the flexibility and reach of the variable region. They also affects the affinity for Fc receptor binding and mediated effector functions [@Kretschmer2017].


## Variability - NOT DONE YET
B cells have several ways to transform their antibodies, be it function or specificity. Function is mainly controlled by the Constant region of the heavy chain, while the variable region of both chains work together to decide what the antibody will attatch and react to.


### Activation-induced cytidine deaminase (AID)
Activation-induced cytidine deaminase is a protein that plays a central role in the adaptive immune system. It plays a central role in the processes of CSR and SHM which are central in the generation of variability. The main mechanism of AID is the controlled deamination of cystein in the variable region of the antibodies. When uracil is detected in DNA it triggers a doublestranded break in the DNA which is then repaired replacing uracil with cystein, as it originally was [@Chi2020]. AID indirectly contribute to the DNA changeing from break to break, by suppressing the normal pathways that the cell would use to repair these kinds of DNA damage. This suppression forces the cell to repair the DNA imperfectly leading to mistakes which contribute to the variation in the cells variable region [@Anvari2018].


#### Somatic Hyper Mutation (SHM)
Somatic hyper mutation is a way for the B cell to insert, delete, and swap bases between generations. It mainly occurs in the variable region of the DNA where it causes minor changes in the DNA making up the variable region, thereby increasing the number of potential combinations that can be created [@Chi2020]. Higher affinity mutations are selected for further mitosis, this causes the next generation of antibody producing B cells to produce antibodies of higher affinity [@Victora2010].


#### Class Switch Recombination (CSR)
Class switch recombination is the process by which B cells change the constant region of the heavy chain of the antibodies they produce. This is done by splicing different parts of the DNA together, cutting out the middle from the splicing [@Liu2024]. This makes it impossible for the transformation to go backwards, as the DNA is not there.
Even though there is no mutation during CSR, AID is necessary for triggering the double strand breaks for recombining the DNA. CSR occurs by breaking two switch regions in the DNA, it is between these two regions that the DNA is removed and recombined to allow for the production of another isotype of antibody [@Liu2024]. With the excision of only some DNA it allows the downstream B cell generations to perform further CSRs for further adaptation.


### V(D)J Recombination
The variability of the variable region comes partially from V(D)J Recombination. V(D)J Recombination is different between the variable regions of the heavy chain and light chain. On the heavy chain it involves all three segments the V(ariable), D(diverse), and J(joining) segments, the light chain only has the V and J segments. By combining multiple different V and J segments it is possible to generate several different combinations [@Chi2020]. By combining these basic forms with inserted bases from SHM it is possible for the immunne system to reach a high level of starting diversity for affinity maturation to improve upon.


## Isotypes
The different isotypes differ structually in various ways, not just in their length but also in their glycosylation and their cystein placement. It is these structural differences that the body uses to selectively interact with the individual antibodies to mediate which reaction is triggered by which isotype [@Kretschmer2017].

### IgM
IgM is the default isotype of mature B cells. It contains 4 conserved regions, and in the blood they form pentamers. They are the best activators of the complement system. IgM can be produced without being exposed to antigenic stimuli, but they can also be produced against invading pathogens [@Kretschmer2017]. It serves as the primary defense against early pathogen invasions, before adaptive immunity has been activated [@Chi2020].

### IgD
IgD is the other type of immunoglobin that is produced by naive B cells [@Stavnezer2008]. It is believed to have arrisen evolutionarily slightly after IgM, and is preserved from fish to humans. The maturation of B cells is characterized by progressive up-regulation of IgD receptors. B cells that purely express IgD are anergic B cells and tend to exist in the perepheri parts of the body, but interestingly not in the intestine [@Gutzeit2018]. This may be due to the presence of the intestinal microfauna. If antibodies were a danger to the commensal microbes there should be a similar situation for IgA antibodies, which are the main excretory antibodies.

### IgG
IgG antibodies come in four subtypes, IgG1, IgG2, IgG3, and IgG4. The subtypes are numberd in order of decreasing abundance. All four subtypes can be transcribed by the G class B cells. The different subtypes have different affinity for different receptors of the Fcγ family. Through the different affinities they fulfil different purposes in the immune system. With the exception of FcγRIIb, all of the FcγR receptors mediate activating functions [@Kapur2014].

#### IgG1
IgG1 is the bodys response to soluble proteins and membrane proteins. They are the most abundant of the IgG antibodies, and are often encountered together with the other IgG subtypes, especially IgG3 and IgG4 [@Vidarsson2014]. IgG1 is able to cross-link and react with all of the different FcγR receptors [@Kapur2014]. 

#### IgG2
IgGs reaction to bacterial capsular polysaccharides is almost completely made up of IgG2 antibodies [@Vidarsson2014]. IgG2 mainly triggers receptors on neutrophiles, it might also be able to trigger receptors on NK cells and some macrophages [@Kapur2014]. This would lower the scope of the reaction, triggering only the necessary cells for fighting the bacterial infection. Therefore no antibodies are wasted on other activities, leading to an improved immune response.

#### IgG3
IgG3 is the antibody with the longest hinge domain, counting up to 62 amino acids in length. It also has the highest affinity for C1q which makes it the best IgG to trigger the complement system. It does have the shortest halflife amongst the long lived IgGs. Just like IgG1 it can interact with all of the FcγR [@Kapur2014].

#### IgG4
IgG4 is an antibody that is often formed following repeated non-infectious exposure. It is linked to downregulation of immune reaction towards the antigen and is usually seen correlating with sucessful immunotherapy [@Vidarsson2014].

### IgA
IgA is the most produced antibody isotype in the body, it is the second most prominent antibody in the blood. More IgA is produced in total than all the other isotypes combined. It is encountered in all mammals and birds. Most IgA is secreted out of the body, usually in the form of SIgA, which forms a dimer outside of the body, and sometimes even a tetramer. IgA comes in two major subtypes, IgA1 and IgA2. An obvious difference between the two subtypes, is the length of their hinge regions. THe significantly longer hinge region of IgA1 makes the arms with the variable region significantly more flexible [@Sousa-Pereira2019].  


### IgE
IgE is the antibody type mainly associated with allergic and parasitic diseases. It mainly interacts with receptors on most cells, basophils, monocytes and macrophages, with some other immune cells carrying lower affinity receptors. When a antibody-receptor complex is triggered the triggerd cells degranulate completely, this degranulatinon triggers high levels of inflammation and cascading reactions [@Kretschmer2017]. The relatively higher affinity of IgE receptors drastically lowers the amount of IgE that is freely found in blood serum.  

IgE is the least abundant antibody in the blood. This is probably partially due to the lower life lifespan when unbound, which is 1.5 days [@Gutzeit2018], and the fact that it is produced in response to parasites, which are more rare than microbial lifeforms. In some individuals IgE antibodies are produced as a response to harmless antigens, this failure can lead to the symptoms of allergy [@Anvari2018]. Perhaps the growing prevalence of allergy is due to the lower variation in IgE leading to an increased amount of Mast cells being triggered by a single harmless antigen.


## Proteomics
Proteomics is the study of a large amount of proteins from the same source. It allows for the answering of questions in relation to the abundance, localisation, interactions, modification states, and folding of proteins. Proteiomics comes in two main types, bottom-up and top-down. In bottom-up proteomics an overall picture of the protein abundance is created and then further studied in order to get an overviw of the situation, and find out how the different proteins play together. Top-down proteomics [@McArdle178].


### Datasource - PRIDE - NOT DONE YET
PRIDE is the online protein repository of the European Molecular Biology Laboratory (EMBL). It stands for The PRoteomics IDEntifications Archive database, and is a public data repository for mass spectrometry data. The goal of PRIDE is to support the performace of reproducible research through the application of quality control metrics and enable the reuse of data for purposes other than the original study. The source of the datasets in this study is PRIDE and each dataset contains the ftp address from which it came [@pride].

### FragPipe - NOT DONE YET
FragPipe is a application designed to be a comprehensive computational platform for the analysis of mass spectrometry proteomics data. It functions as a graphical wrapper over several proteomics tools, such as MSFragger, IonQuant, DiaNN, and more. It provides a comprehensive list of workflows that combine the tools to turn multistep processes into a set up and run. FragPipe is open source, and under continuous development [@FragPipe].


# Project Description

## Introduction and Background
Antibodies are proteins that bind antigens, in humans they come in 5 different isotypes. Each isotype fills a different role in the immune system. As different isotypes fulfil different roles in the immune system, different diseases should have different isotype fingerprint.

It is expected that different categories of diseases with different attack mechanisms will trigger different reactions from the immune system, including provoking the creation of different isotype fingerprints. These fingerprints should be able to tell us about the disease, and how the body fights it. By understanding the body’s general reaction it will be possible to plan personalized treatment.

With the help of proteomic methodology, it is possible to map the relative concentrations of antibodies from blood samples. By modifying proteomics data analysis workflows like FragPipe to run with specifically designed databases, it will be possible to measure the relative antibody isotype distributions and provide in-depth antibody profiling. 

A main case to look into are the differences between the isotype profiles of allergies (hyperimmune reactions) where there is expected to be an increase in IgE, and the profile of a bacterial infection, which is expected to have a higher concentration of IgG. If the general fingerprint of the disease can be well characterized via analysis of public proteomics data, this will open the door for diagnosis via antibody profiling with high analytic depth coming from the high sensitivity of common bottom-up mass spectrometry experiments. .

This project will build on a prior ISA project where we investigated antibody isotype distributions in genomic data, and found a marked increase in presence of IgD producing B-cells in allergic reactions, which was more noticeable than the increase in IgE, which should be the main reaction. We therefore see the need to characterize these patterns directly on protein level, thus on the presence of actually expressed antibodies.



## Aim of the Project
The aim of the project is to perform peptide analysis on databases to compare isotype fingerprint between different categories of diseases, such as viral, autoimmune, and whatever else can be found. Healthy individuals will be used as a control group.


## Methods and Experimental Design
We will check repositories like PRIDE to see if they contain any mass spectra that are suitable for the project, such as serum measurements of sick patients. We will re analyse the identified raw data using fragpipe, with a focus on antibody isotypes. The goal is to create an overview of isotype distribution for further analysis.

After gathering the isotype data, we will profile the databases and perform statistical comparison between the different diseases. Theoretically the potentially high variability of isotypes will average out on isotype level by increasing the sample size.

If possible we will then also look for disease specific biomarker peptides that could affect the antibody profile. Once that has been done we can train some simple machine learning models, such as random forest.

Antibody profiling will allow us to better understand the immune response to diseases and potentially identify new mechanisms for future treatments. Another potential end goal would be the optimization of vaccine adjuvants to design the optimal antibody profile with the vaccines, to avoid allergic reactions and stimulate the generation of the most effective isotype profiles. 


# Method - NOT DONE YET
## Data Collection
Data was gathered from the pride repository. The goal was to gather datasets of serum samples, from as many differnet diseases as possible. First ChatGPT was asked for serum samples from various diseases, it forgot partway through the conversation that it was supposed to give serum samples, and it wasn't very thorough with the list it created. After having processed the datasets that the chatbot had found, another attempt was made with claude, which produced a python code for questioning the pride api, plus keywords. Finally a brute force search was conducted going through all entries on pride that had organism "homo sapien" and organism part as one of the following; "blood serum, blood, or blood plasma.

The next step was to look through the article related to the datasets of interest to find the parameters to use for FragPipe analysis, and to see if there were any issues with the dataset in relation to the project itself. This aimed to remove the datasets where there were issue for the study. The datasets where there was a need to guess which dataset belonged to control and which was of interest. THe datasets where antibodies were removed fo increase the clarity of other proteins. The datasets that were targeting something even more specific in the blood or serum.


## Data Treatment
After the datasets were verified, the few that had all the information needed, a proper separation of samples noted, without any antibodies removed, and not targeting something irrelevant in the blood, were ran through FragPipe.


## Data Storage
The rows of interest were extracted with csvgrep (from csvkit) and aggregated (with csvstack from csvkit) the different patients into a single csv, with an new column called patient_id. After that the aggregated csv was input into postgres by using the csvsql function of csvkit.

Another datatable with patient information was established inside the same database. The data source can be found further down the article under the Data section.


## Data Visualization
Data visualization was performed with R using the ggplot2 package. The data was normalized and then compared in categories. The goal was originally to compare the distribution of isotypes accross different disease categories, but due to a lack of usable data some compromises have been made.




\begin{center}
\tikzset{
  block/.style = {
    rectangle, 
    draw, 
    text width=10em, 
    text centered, 
    rounded corners, 
    minimum height=3em
  }
}

\begin{tikzpicture}
[node distance = 5cm,
start chain=going below,]
\node (n1) at (0,0) [block] {\includegraphics*[height=2.5em]{PRIDE}};
\node (n2) [block, right of=n1] {\includegraphics*[height=2.5em]{FragPipe}};
\node (n3) [block, right of=n2] {\includegraphics*[height=2.5em]{Csvkit}};
\node (n4) [block, below of=n1, yshift =2cm] {\includegraphics*[height=2.5em]{PostgreSQL}};
\node (n5) [block, right of=n4] {\includegraphics*[height=2.5em]{DBI}};
\node (n6) [block, right of=n5] {\includegraphics*[height=2.5em]{R}};

\draw [->] (n1) -- (n2);
\draw [->] (n2) -- (n3);
\draw [thick,->] (n3) -- (n4);
\draw [->] (n4) -- (n5);
\draw [->] (n5) -- (n6);
\end{tikzpicture}
\end{center}

```{r}
#| echo: false
#| fontsize: \tiny
library(DBI)
library(RPostgres)
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
SIG_FIG <- 2
con <- dbConnect(RPostgres::Postgres(), dbname = "antibodies")

```


# Results
```{r}
#| echo: true
# select patient, protein_id, gene, total_peptides from proteins_small;
peptides <- dbGetQuery(con, "select patients.patient_id, patients.disease_type, patients.disease, 
                       proteins_small.protein_id, proteins_small.gene, 
                       proteins_small.total_peptides
                       from proteins_small
                       left join patients
                       on proteins_small.patient = patients.patient_id")
head(peptides, 20)
peptides_mutated <- peptides |>
  pivot_wider(
    id_cols = c(patient_id, disease_type, disease),
    names_from = gene,
    values_from = total_peptides
  ) |>
  mutate(
    total = rowSums(across(starts_with("igh")), na.rm = TRUE),
    pct_ighg1 = round(ighg1 / total * 100, SIG_FIG),
    pct_ighg2 = round(ighg2 / total * 100, SIG_FIG),
    pct_ighg3 = round(ighg3 / total * 100, SIG_FIG),
    pct_ighg4 = round(ighg4 / total * 100, SIG_FIG),
    pct_ighm = round(ighm / total * 100, SIG_FIG),
    pct_igha1 = round(igha1 / total * 100, SIG_FIG),
    pct_igha2 = round(igha2 / total * 100, SIG_FIG),
    pct_ighd = round(ighd / total * 100, SIG_FIG),
    total_pct = rowSums(across(starts_with("pct_igh")), na.rm = TRUE)
  ) |>
  pivot_longer(
    !c(patient_id, disease_type, disease),
    names_to = "isotype",
    values_to = "count"
  )
# write.csv(peptides_wide, file = "peptides_wide.csv")
peptides_pct <- peptides_mutated[grepl("pct_i", peptides_mutated$isotype),]

head(peptides_pct, 20)

ggplot(peptides_pct, aes(x = disease, y = count, color = isotype)) +
  geom_boxplot()


peptides_collapse <- peptides_pct |> # gather all the different ighg and igha into just ighg and igha
  mutate(isotype = if_else(str_detect(isotype, "ighg"), "pct_ighg", isotype)) |>
  mutate(isotype = if_else(str_detect(isotype, "igha"), "pct_igha", isotype)) |>
  group_by(patient_id,disease_type, disease, isotype) |>
  summarise(count = sum(count), .groups = "drop")

head(peptides_collapse, 10)

ggplot(peptides_collapse, aes(x = disease, y = count, color = isotype)) +
  geom_boxplot() 
```

```{r}
stats_pct <- peptides_pct |> # find the mean and standard deviation for each disease
  group_by(disease_type, disease, isotype) |> 
  summarise(mean = mean(count, na.rm = TRUE), std = sd(count, na.rm = TRUE))

head(stats_pct)

ggplot(stats_pct, aes(x = disease, y = mean, color = isotype)) +
  geom_point() +
  labs(title = "Mean of all subtypes")

ggplot(stats_pct, aes(x = isotype, y = mean, color = disease)) +
  geom_point() +
  labs(title = "Mean of all subtypes")
```



```{r}
stats_collapse <- peptides_collapse |> # find the mean and standard deviation for each disease
  group_by(disease_type, disease, isotype) |>
  summarise(num = length(count), mean = mean(count, na.rm = TRUE), std = sd(count, na.rm = TRUE))

head(stats_collapse)

ggplot(stats_collapse, aes(x = disease, y = mean, color = isotype)) +
  geom_point() +
  labs(title = "Mean of collapsed isotypes")

ggplot(stats_collapse, aes(x = isotype, y = mean, color = disease)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(title = "Mean of collapsed isotypes")
```

```{r}
print(stats_collapse[order(stats_collapse$isotype, decreasing = TRUE),])

#lm(peptides_collapse)
```
# Discussion


# Conclusion

# Data 

```{r}
#| echo: false
#| fontsize: \tiny
data <- dbReadTable(con, "patients")
dataset <- rep("PXD036074", 31)
ftp <- c("20220624_Z1_ZW_001_30-0066_Ctr-A1_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A2_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A3_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A4_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A5_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A6_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A7_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A8_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A9_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A10_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A11_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-A12_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-B1_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-B2_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_Ctr-B3_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E1_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E2_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E3_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E4_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E5_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E6_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E7_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E8_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E9_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_COVID-E10_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_MPX-D1_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_MPX-D2_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_MPX-D3_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_MPX-D4_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_MPX-D5_heat_DIA.wiff2.Zip",
  "20220624_Z1_ZW_001_30-0066_MPX-D6_heat_DIA.wiff2.Zip")
overview <- cbind(data[,4],dataset,ftp)
options(
  pillar.align = "left",
  pillar.min_title_chars = 0)
print(as.data.frame(overview))
dbDisconnect(con)
```

# Bibliography
